A Reimplementation of Python’s dict using Simple Tabulation Hashing and Linear Probing

### 2012-05-20 docs/paper.tex

\title{A Reimplementation of Python’s dict using Simple Tabulation Hashing and Linear Probing}
\author{
        Russell Cohen \\
            \and
        Thomas Georgiou \\
            \and
        Pedram Razavi
}
\date{\today}
\documentclass[11pt]{article}
\usepackage[pdftex]{graphicx}
\usepackage{setspace}
\usepackage{fullpage}

@@

For instance, a dictionary can be instantiated to pass keyword arguments to a function. This means a dictionary could potentially be created and destroyed on every function call.  Some other examples of internal usage of dictionaries in Python are as follows:
\begin{itemize}
\item Class method lookup
\item Instance attribute lookup and global variables lookup
\item Python built-ins module
\item Membership testing
\end{itemize}

@@

In this section we give a brief overview of the key implementation details of dictionaries in CPython 2.7.3.
We should note that Python’s dictionary supports several different data types as keys. However as we noted earlier, the dictionaries underlying class instances and modules have only strings as keys. Therefore optimizing a dictionary which have string-only keys is crucial to the runtime of any Python program. CPython accommodates for this optimization by changing its dictionary lookup method as necessary. First when a dictionary is instantiated, CPython uses a string-only lookup function until a search for non-string is requested, in this case CPython falls back to a more general lookup method and uses this general function onward. There are two main optimizations in the string-only lookup function. First, since the string comparisons do not raise any exceptions by design, some of the error checking logic can been removed. Second, the string-only lookup function can skip the rich set of comparisons ( <=, >=, <, >, ==, !=) which arbitrary data types can provide since string type does not have these special cases.  As it can be seen, string-only keys can dominate the runtime of a dictionary and CPython uses some measures to optimize for this common case. Because of this importance we mostly limit ourselves to the study of new implementations of hashing the string type for the rest of the paper.

@@

On the other hand this hashing strategy has the tendency to fill contiguous slots of the
hash table. CPython tries to solve this behavior by utilizing a good collision resolution strategy. The collision resolution strategy currently used is open addressing with a custom non-linear probing scheme. In this scheme CPython uses quadratic probing based on this function: $j = (5j + 1) \bmod 2^i$ where $0<j<2^i-1$. This recurrence alone is not enough for a better behavior since again it behaves the same as linear-probing for some keys since it scans the table entries in a fixed order. To overcome this issue CPython also uses the most significant bits of the hash value with the following code snippet:
\begin{verbatim}
slot = hash;
perturb = hash;
while (<slot is full> && <item in slot does not equal the key>) {
       slot = (5*slot) + 1 + perturb;
       perturb >>= 5;
}
\end{verbatim}
Based on this new method the probe sequence will depend on nearly all the bits of a hash value. We should note that the value of  perturb shift (currently 5) is crucial for a good probing behavior. Since it should be small enough so that  all the significant bits have an  effect on the probe sequence but at the same time it should be large enough so that in really bad cases the high-order hash bits can affect the earlier iterations.

>>>>

\title{A Reimplementation of Python’s dict using Simple Tabulation Hashing and Linear Probing}
\author{
        Russell Cohen \\
            \and
        Thomas Georgiou \\
            \and
        Pedram Razavi
}
\date{\today}
\documentclass[11pt]{article}
\usepackage[pdftex]{graphicx}
\usepackage{setspace}
\usepackage{fullpage}
\usepackage{float}

@@

For instance, a dictionary can be instantiated to pass keyword arguments to a
function. This means a dictionary could potentially be created and destroyed on
every function call.  Some other examples of internal usage of dictionaries in
Python are as follows:

\begin{itemize}
\item Class method lookup
\item Instance attribute lookup and global variables lookup
\item Python built-ins module
\item Membership testing
\end{itemize}

@@

In this section we give a brief overview of the key implementation details of
dictionaries in CPython 2.7.3.  We should note that Python’s dictionary supports
several different data types as keys. However as we noted earlier, the
dictionaries underlying class instances and modules have only strings as keys.
Therefore optimizing a dictionary which have string-only keys is crucial to the
runtime of any Python program. CPython accommodates for this optimization by
changing its dictionary lookup method as necessary. First when a dictionary is
instantiated, CPython uses a string-only lookup function until a search for
non-string is requested, in this case CPython falls back to a more general
lookup method and uses this general function onward. There are two main
optimizations in the string-only lookup function.  First, since the string
comparisons do not raise any exceptions by design, some of the error checking
logic can been removed. Second, the string-only lookup function can skip the
rich set of comparisons ( $<=$, $>=$, $<$, $>$, $==$, $!=$) which arbitrary data
types can provide since string type does not have these special cases.  As it
can be seen, string-only keys can dominate the runtime of a dictionary and
CPython uses some measures to optimize for this common case. Because of this
importance we mostly limit ourselves to the study of new implementations of
hashing the string type for the rest of the paper.

@@

On the other hand this hashing strategy has the tendency to fill contiguous
slots of the hash table. CPython tries to solve this behavior by utilizing a
good collision resolution strategy. The collision resolution strategy currently
used is open addressing with a custom non-linear probing scheme. In this scheme
CPython uses quadratic probing based on this function: $j = (5j + 1) \bmod 2^i$
where $0<j<2^i-1$. This recurrence alone is not enough for a better behavior
since again it behaves the same as linear-probing for some keys since it scans
the table entries in a fixed order. To overcome this issue CPython also uses the
most significant bits of the hash value with the following code snippet:

\begin{verbatim}
slot = hash;
perturb = hash;
while (<slot is full> && <item in slot does not equal the key>) {
       slot = (5*slot) + 1 + perturb;
       perturb >>= 5;
}
\end{verbatim}

Based on this new method the probe sequence will depend on nearly all the bits
of a hash value. We should note that the value of  perturb shift (currently 5)
is crucial for a good probing behavior. Since it should be small enough so that
all the significant bits have an  effect on the probe sequence but at the same
time it should be large enough so that in really bad cases the high-order hash
bits can affect the earlier iterations.
